---
title: "04-world-series-home-field"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
require(dplyr)
require(data.table)
library(readxl)
```

## Probability and Inference --- Deliverable 04
### World Series Home Field
### Rastko Stojsin

## Setup
In the last deliverable I looked at a teams probability of winning a series given the probability of winning a single game and length of series. In the real world the probability of winning a single game is dependent on many outside factors. A big factor, especially in baseball, is home field advantage. Here I will analyse how much of a role home field advantage has in baseball in determining a series outcome. I will only look at 7 game series between two baseball teams.

In a 7 game baseball series the possible schedules are shown in the table below.

 Overall Advantage | Game 1 | Game 2 | Game 3 | Game 4 | Game 5 | Game 6 | Game 7 
|-|-|-|-|-|-|-|-|
| $Team A$ | $Stadium A$ | $Stadium A$ | $Stadium B$ | $Stadium B$ | $Stadium B$ | $Stadium A$ | $Stadium A$ |
| $Team B$ | $Stadium B$ | $Stadium B$ | $Stadium A$ | $Stadium A$ | $Stadium A$ | $Stadium B$ | $Stadium B$ |

Next lets say that playing at home gives a team a 10% (P(winning single head to head game) * 1.1) bump in winning percentage.

## Questions

### Q1
1. Compute analytically the probability that the Team A win the world series when the sequence of game locations is {Home, Home, Away, Away, Away, Home, Home}. Calculate the probability with and without home field advantage when P1 = 0.55. What is the difference in probabilities?

```{r load_excel, include=FALSE}
# Get all possible outcomes
all_possibilities <- read_excel("../all-possible-world-series-outcomes.xlsx")
# set all posibilitys to a data table
all_possibilities <- as.data.table(all_possibilities)
# Home field indicator
```

```{r, set_variables, include=FALSE}
# set inital variables
home_field_indicator <- c(1,1,0,0,0,1,1) #{h, h, a, a, a, h, h}
prob_single_game_win <- 0.55
advantage_multiplier <- 1.1 # Set = 1 for no advantage
prob_home_win <- prob_single_game_win*advantage_multiplier
prob_away_win <- 1 - (1 - prob_single_game_win)*advantage_multiplier
```

```{r, calculate_probabilitys, include=FALSE}
# Calculate the probability of each possible outcome
all_possibilities[, p := NA_real_] # Initialize new column in apo to store prob
for(i in 1:nrow(all_possibilities)){
  prob_series <- rep(1, 7)
  for(j in 1:7){
    p_win <- ifelse(home_field_indicator[j], prob_home_win, prob_away_win)
    prob_series[j] <- case_when(
        all_possibilities[i,j,with=FALSE] == "W" ~ p_win
      , all_possibilities[i,j,with=FALSE] == "L" ~ 1 - p_win
      , TRUE ~ 1
    )
  }
  all_possibilities[i, p := prod(prob_series)] # Data.table syntax
}
```

```{r output_with_homefield_adv}
# Probability of overall World Series outcomes with homefield advantage 
win_series_with_homefield <- all_possibilities[, sum(p), overall_outcome]
win_series_with_homefield
```

We can see that with homefield advantage in play Team A would win the world series with a probability of around 0.6345 given they win 0.55 normally with a 1.1 boost at home and a 1.1 boost for team B at their home and a series schedule of {Home, Home, Away, Away, Away, Home, Home}. Lets see how this compaires to if homefield advantage did not exist.

```{r, calculate_probabilitys_no_homefield_adv, include=FALSE}
# Calculate the probability of each possible outcome
all_possibilities[, p := NA_real_] # Initialize new column in apo to store prob
for(i in 1:nrow(all_possibilities)){
  prob_series <- rep(1, 7)
  for(j in 1:7){
    p_win <- ifelse(home_field_indicator[j], 0.55, 0.55)
    prob_series[j] <- case_when(
        all_possibilities[i,j,with=FALSE] == "W" ~ p_win
      , all_possibilities[i,j,with=FALSE] == "L" ~ 1 - p_win
      , TRUE ~ 1
    )
  }
  all_possibilities[i, p := prod(prob_series)] # Data.table syntax
}
```

```{r output_without_homefield_adv}
# Probability of overall World Series outcomes without homefield advantage 
win_series_without_homefield <- all_possibilities[, sum(p), overall_outcome]
win_series_without_homefield
# Difference in World Series outcome between with and without homefield advantage
win_series_with_homefield$V1[1] - win_series_without_homefield$V1[1]
```

If there is no homefield advantage the probability of team A winning the same series drops by 0.0262 to 0.6082.

### Q2

2. Calculate the same probabilities as the previous question by simulation.

```{r set_vars_&_make_function, include=FALSE}
# set home field indicators
hfi1 <- c(1,1,0,0,0,1,1) #{ATL, ATL, NYC, NYC, NYC, ATL, ATL}
hfi2 <- c(0,0,1,1,1,0,0) #{ATL, ATL, NYC, NYC, NYC, ATL, ATL}

# build function
world_series_win <- function(hfi, p_win_game, advan_multip){
  wins <- 0
  games <- 0
  for(i in 1:7){
    if(hfi[i]){
      p_game <- p_win_game* advan_multip
    }else{
      p_game <- (1-p_win_game) * advan_multip
    }
    game_outcome <- rbinom(1,1,p_game)
    wins <- wins + game_outcome
    games <- games + 1
    if(wins == 4 | games - wins == 4) break
  }
  return(wins>3)
}
```

```{r run_sim, include=FALSE}
set.seed(537)
sims <- 1000000
hfi_vec_g1 <- sample("hfi1", sims, replace = TRUE)
hfi_vec_g2 <- sample("hfi2", sims, replace = TRUE)

num.seq <- sample(1:sims, 1)
team1.win.g1 <- rep(NA, sims)
team1.win.g2 <- rep(NA, sims)
team1.win.g3 <- rep(NA, sims)

for(i in 1:sims){
  team1.win.g1[i] <- world_series_win(hfi = hfi1, 0.55, 1.1)
  team1.win.g2[i] <- world_series_win(hfi = hfi2, 0.55, 1.1)
 }
```

```{r sim_results}
# win percentage with homefield advantage and with (h,h,a,a,a,h,h) schedule
win_p_sim_g1 <- mean(team1.win.g1)
win_p_sim_g1
# win percentage with out homefield advantage
win_p_sim_witoutHA <- mean(team1.win.g2)
win_p_sim_witoutHA
```

### Q3

3. What is the absolute and relative error for your simulation in the previous question?

```{r}
# absolute error win with homefield advantage
abs_wA <- abs(win_series_with_homefield$V1[1] - win_p_sim_g1)
abs_wA
# absolute error win without homefield advantage
abs_woA <- abs(win_series_without_homefield$V1[1] - win_p_sim_witoutHA)
abs_woA
# relative error win with homefield advantage
rel_wA <- abs_wA/win_series_with_homefield$V1[1]
rel_wA
# relative error win without homefield advantage
rel_woA <- abs_woA/win_series_without_homefield$V1[1]
rel_woA
```

### Q Bonus
Bonus: Does the difference in probabilites (with vs without home field advantage) depend on P team 1 wins?
```{r, include=FALSE}
home_field_indicator <- c(1,1,0,0,0,1,1) #{h, h, a, a, a, h, h}
prob_single_game_win <- 0.70
advantage_multiplier <- 1.1 # Set = 1 for no advantage
prob_home_win <- prob_single_game_win*advantage_multiplier
prob_away_win <- 1 - (1 - prob_single_game_win)*advantage_multiplier
all_possibilities[, p := NA_real_] # Initialize new column in apo to store prob
for(i in 1:nrow(all_possibilities)){
  prob_series <- rep(1, 7)
  for(j in 1:7){
    p_win <- ifelse(home_field_indicator[j], prob_home_win, prob_away_win)
    prob_series[j] <- case_when(
        all_possibilities[i,j,with=FALSE] == "W" ~ p_win
      , all_possibilities[i,j,with=FALSE] == "L" ~ 1 - p_win
      , TRUE ~ 1
    )
  }
  all_possibilities[i, p := prod(prob_series)] # Data.table syntax
}
win_series_with_homefield_70 <- all_possibilities[, sum(p), overall_outcome]
win_series_with_homefield_70
all_possibilities[, p := NA_real_] # Initialize new column in apo to store prob
for(i in 1:nrow(all_possibilities)){
  prob_series <- rep(1, 7)
  for(j in 1:7){
    p_win <- ifelse(home_field_indicator[j], 0.70, 0.70)
    prob_series[j] <- case_when(
        all_possibilities[i,j,with=FALSE] == "W" ~ p_win
      , all_possibilities[i,j,with=FALSE] == "L" ~ 1 - p_win
      , TRUE ~ 1
    )
  }
  all_possibilities[i, p := prod(prob_series)] # Data.table syntax
}
# Probability of overall World Series outcomes without homefield advantage 
win_series_without_homefield_70 <- all_possibilities[, sum(p), overall_outcome]
win_series_without_homefield_70
```
```{r}
# difference of homefield advantage with team 1 winning 0.55
win_series_with_homefield$V1[1] - win_series_without_homefield$V1[1]
# difference of homefield advantage with team 1 winning 0.70
win_series_with_homefield_70$V1[1] - win_series_without_homefield_70$V1[1]
```

Yes it does! The importance of homefield advantage rises as the probability of team 1 winning an individual game goes up. This makes logical sence as well, as homefield advantage isnt a set amount (ie raises probability by 0.05), it is in fact a ratio based on the individual game win percentage. It raises the individual win percentage by 10%, and 10% of x>0.55 is larger than 10% of 0.55 and 10% of x<0.55 is smaller!

### Q Bonus

Bonus: Does the difference in probabilites (with vs without home field advantage) depend on the advantage factor? (The advantage factor in PH and PA is the 1.1 multiplier that results in a 10% increase for the home team.)

```{r , include=FALSE}
home_field_indicator <- c(1,1,0,0,0,1,1) #{h, h, a, a, a, h, h}
prob_single_game_win <- 0.55
advantage_multiplier <- 1.7 # Set = 1 for no advantage
prob_home_win <- prob_single_game_win*advantage_multiplier
prob_away_win <- 1 - (1 - prob_single_game_win)*advantage_multiplier
all_possibilities[, p := NA_real_] # Initialize new column in apo to store prob
for(i in 1:nrow(all_possibilities)){
  prob_series <- rep(1, 7)
  for(j in 1:7){
    p_win <- ifelse(home_field_indicator[j], prob_home_win, prob_away_win)
    prob_series[j] <- case_when(
        all_possibilities[i,j,with=FALSE] == "W" ~ p_win
      , all_possibilities[i,j,with=FALSE] == "L" ~ 1 - p_win
      , TRUE ~ 1
    )
  }
  all_possibilities[i, p := prod(prob_series)] # Data.table syntax
}
win_series_with_homefield_1.7 <- all_possibilities[, sum(p), overall_outcome]
win_series_with_homefield_1.7
all_possibilities[, p := NA_real_] # Initialize new column in apo to store prob
for(i in 1:nrow(all_possibilities)){
  prob_series <- rep(1, 7)
  for(j in 1:7){
    p_win <- ifelse(home_field_indicator[j], 0.55, 0.55)
    prob_series[j] <- case_when(
        all_possibilities[i,j,with=FALSE] == "W" ~ p_win
      , all_possibilities[i,j,with=FALSE] == "L" ~ 1 - p_win
      , TRUE ~ 1
    )
  }
  all_possibilities[i, p := prod(prob_series)] # Data.table syntax
}
# Probability of overall World Series outcomes without homefield advantage 
win_series_without_homefield_1.7 <- all_possibilities[, sum(p), overall_outcome]
win_series_without_homefield_1.7
```

```{r}
# difference of homefield advantage with team 1 winning 0.55 and a homefield advantage of 10%
win_series_with_homefield$V1[1] - win_series_without_homefield$V1[1]
# difference of homefield advantage with team 1 winning 0.55 and a homefield advantage of 70%
win_series_with_homefield_1.7$V1[1] - win_series_without_homefield_1.7$V1[1]
```

Yes it does! The difference in probabilities (with/without homefield advantage) is heavily reliant on the level of advantage. The higher advantage level the higher the difference between advantage and no advantage.

